Scalable statistical genetic data analysis in Python

### 2024-01-09 paper.tex
Initial pass at distributed compute narrative

# Corresponding shell
tail -n+2 sample_info.csv | awk '$2 >= 1950'
    | cut -f1 > samples.txt
bcftools view PATH -s samples.txt
    | bcftools +fill-tags | bcftools +af-dist

\end{minted}
\end{tcolorbox}
\end{tabular}
\caption{Compute performance on subsets of the matrix.
(A) Total CPU time required to run the afdist calculation for
a subset of 10 samples and 10000 variants from the middle of the matrix.
(B) Same for n/2 samples. Genozip did not run in this case for
$n > 10^4$ samples because it does not support a file to specify
sample IDs, and the command line was therefore too long for the shell
to execute.
(C) A simplified version of the source code used for these benchmarks.
\label{fig-subset-matrix-compute}}
\end{figure*}

\subsection{Distributed computation}
Why do we need distributed compute? Summary of current
approaches (chunk up along the genome using HPC or workflow
managers). Advantages of true distributed compute, MapReduce
etc. Summary of earlier efforts like ADAM.
Discuss Hail. It's awesome, has enabled current
generation of mega-scale projects. Drawbacks are the
architecture has emphasised performance etc over adaptability
and simplicity. File format is undocumented, etc.
Discuss why the sgkit architecture is suited to distributed
computation. Discuss Zarr's use in cloud stores, and existing
deployments. Discuss Dask.
We do not attempt a head-to-head performance comparison with
Hail here, but defer to Liangde's thesis~\citep{li2022efficient}
with brief summary.
Forward ref to following sections on applications to demonstate
sgkit in a distributed context.

@@

\subsection{Statistical Genetics}
Statistical genetics is largely concerned with disease associations
in humans, and the success of genome wide association studies (GWAS) has led
to explosive growth in the field~\citep{uffelmann2021genome}.
Sample size is the single most
important factor in driving statistical power, and
genotype array data has been collected for millions of samples
[more explanation and citations].
GWAS typically use genotype array data because of
its continuing cost-effectiveness, although there has been
exome~\citep{lek2016analysis,backman2021exome}
and large-scale whole genome sequence data is becoming increasingly
available~\citep{halldorsson2022sequences}.
Genotype array data is usually imputed using software such
as BEAGLE~\citep{browning2018one} or IMPUTE~\citep{howie2011genotype}.
% Mention phasing?

@@

\subsubsection{Distributed computing}

Beyond a certain scale of data there is no alternative but to
parallelise calculations across multiple computers if you wish
them to complete in a reasonable timeframe. Classically in
High Performance Computing (HPC) environments, most often used
to perform scientific computing, such distributed computing
is done using the Message Passing Interface (MPI). MPI
typically assumes a low-latency connection between nodes in
cluster, often with specialised hardware interconnects,
and can be used very effectively to [do things like solve big PDEs].
However, MPI is [hard to program and hard to run], and applications
in genomics have been limited to [things like the ABySS assembler].

[Paragraph describing, Big data, Hadoop, MapReduce, Spark, etc. Why
they don't use MPI, that the problem being solved was/is.
Applications to genomics like ADAM~\citep{nothaft2015rethinking} were promising, but never really
took off. Hail (see also genomics section) notably does a
bunch of things well, but only by rewriting whole chunks of the
underlying technologies]

We use Dask [for reasons].

>>>>

# Corresponding shell
% tail -n+2 sample_info.csv | awk '$2 >= 1950'
    | cut -f1 > samples.txt
bcftools view PATH -s samples.txt
    | bcftools +fill-tags | bcftools +af-dist
\end{minted}
\end{tcolorbox}
\end{tabular}
\caption{Compute performance on subsets of the matrix.
(A) Total CPU time required to run the afdist calculation for
a subset of 10 samples and 10000 variants from the middle of the matrix.
(B) Same for n/2 samples. Genozip did not run in this case for
$n > 10^4$ samples because it does not support a file to specify
sample IDs, and the command line was therefore too long for the shell
to execute.
(C) A simplified version of the source code used for these benchmarks.
\label{fig-subset-matrix-compute}}
\end{figure*}

\subsection{Distributed computation}
Distributed computation has been the reality in the analysis
of genetic variation for many years,
where data is most often provided as per-chromosome VCFs.
This chunking provides a fairly straightforward way to parallelise
computations, using HPC schedulers to split work per-file.
When more fine-grained parallelism is required, users must
provide exact genome coordinates of the genome regions
as part of their submission scripts. Combining results
is a manual operation.
Workflow engines such as
Snakemake~\cite{koster2012snakemake,molder2021sustainable}
and [TODO mention WDL, Cromwell, etc?] make such analyses
much easier, but there is still a great deal of room for
mistakes.
Per-chromosome VCFs are not practical for biobank scale
projects, however, and variant call data tends to be split
into many (still large) files.
For example, the VCFs for 150K UKB WGS data~\cite{halldorsson2022sequences}
are provided in 50kb chunks~\cite{browning2023statistical}, resulting in
hundreds of files per chromosome.
% This is horrible writing, but the point is important.
While these files provide a natural
unit for distributing work, the details of how they
are split differ and essentially there is
no interoperability among large-scale datasets because the VCFs are
so cumbersome that a separate orchestration layer is required to
access the data in parallel.

Despite this long history of de-facto distributed computation
and the clear need for better methods of accessing genetic
variation data in a distributed fashion, there is almost
no scientific literature on the topic (in stark contrast to the
study of methods to compress a VCF into a single file,
as discussed in section XXX).

[This is a really rough outline, also not sure how exactly
we sequence the ideas here. Just getting paragraph fragments
on paper. We do need to explain something of the hadoop ecosystem
to readers who are not familiar, though.]
Distributed analytics for large-scale data is a mature
techology, with many different approaches. Hadoop~\citep{white2012hadoop}
is the original method, and remains the basis of the
dominant ecosystem. HDFS. Spark [CITATION?] lets us
do more complex analytics. Parquet [CITATION?]

Although the Hadoop ecosystem has been hugely successful
in many domains, it has had a limited impact in genomics.
[List some things that have been tried for e.g variant calling
, like ADAM~\citep{nothaft2015rethinking}].
[Briefly mention two approaches using Hadoop clusters
~\citep{boufea2017managing,fan2020variant}, to not very
exiting results].

[Hadoop/Spark/Parquet are fundamentally unsuited to genetic
variation data [WHY]?]

Discuss Hail. It's awesome, has enabled current
generation of mega-scale projects. Drawbacks are the
architecture has emphasised performance etc over adaptability
and simplicity. Uses bits of Spark for X, but custom YZ .
File format is undocumented, etc.

Discuss why the sgkit architecture is suited to distributed
computation. Discuss Zarr's use in cloud stores, and existing
deployments. Discuss Dask.
We do not attempt a head-to-head performance comparison with
Hail here, but defer to Liangde's thesis~\citep{li2022efficient}
with brief summary.
Forward ref to following sections on applications to demonstate
sgkit in a distributed context.

@@

\subsection{Statistical Genetics}
Statistical genetics is largely concerned with disease associations
in humans, and the success of genome wide association studies (GWAS) has led
to explosive growth in the field~\citep{uffelmann2021genome}.
Sample size is the single most
important factor in driving statistical power, and
genotype array data has been collected for millions of samples
[more explanation and citations].
GWAS typically use genotype array data because of
its continuing cost-effectiveness, although there has been
exome~\citep{lek2016analysis,backman2021exome}
and large-scale whole genome sequence data is becoming increasingly
available~\citep{halldorsson2022sequences,uk2023whole}.
Genotype array data is usually imputed using software such
as BEAGLE~\citep{browning2018one} or IMPUTE~\citep{howie2011genotype}.
% Mention phasing?

@@
